Evolutionary rescue

### 2022-10-12 EvolutionaryRescue1.tex
Add files via upload

\section{Model}
We model evolutionary rescue of cancer cells through aneuploidy with the help of multitype branching processes: we have three populations of cancer cells, wildtype euploid $w_t$, wildtype aneuploid $a_t$ and mutant euploid cells $m_t$ which proliferate and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively. Wildtype cells can acquire aneuploidy at rate $u$ while aneuploidy and wildtype cells can mutate with rate $v$ (se Figure \ref{figureAneuploidy}). Changes in the number of wildtype euploid, wildtype aneuploid and mutant euploid cells occur at rates:
\begin{align*}
w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
w_t&\rightarrow w_t-1:\quad \left(\mu_w+u\right)w_t,\\
a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
a_t&\rightarrow a_t-1:\quad \left(\mu_a-w\right)a_t,\\
m_t&\rightarrow m_t+1:\quad \lambda_am_t+wa_t,\\
m_t&\rightarrow m_t-1:\quad \mu_am_t.
\end{align*}
The survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation:
\begin{align}\nonumber
1-p_w&=\frac{\mu_w}{\lambda_w+\mu_w+u+v}+\frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right)+\frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2\\ \label{quadraticeqev1}
&+\frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right),
\end{align}
where the $p_a$ and $p_w$ are the survival probabilities of a population consisting initially of one aneuploidy and wildtype cell, respectively.

 The smallest solution of the quadratic equation \eqref{quadraticeqev1} gives the survival probability:
\begin{align}\label{survprobw}
p_w=\frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w}.
\end{align}
If the population originally consists of a single aneuploidy cell, then the probability that the population will survive is given by the following quadratic equation:
\begin{align}
1-p_a=\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2,
\end{align}
whose smallest solution is the survival probability:
\begin{align}\label{survproba}
p_a=\frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}.
\end{align}
The probability that a single mutant cell will survive is given by \cite{allen2010introduction}:
\begin{equation}
p_m=\left\{
  \begin{array}{@{}ll@{}}
  \frac{\lambda_m-\mu_m}{\lambda_m},\quad\text{if }\lambda_m>\mu_m \\
   0,\quad\text{else}.
  \end{array}\right.
  \end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\begin{tikzpicture}[very thick,decoration={
    markings,
    mark=at position 0.5 with {\arrow{>}}}, middlearrow/.style 2 args={
        decoration={             
            markings, 
            mark=at position 0.5 with {\arrow[xshift=3.333pt]{triangle 45}, \node[#1] {#2};}
        },
        postaction={decorate}
    }]
\edef\x{6.5} 
\edef\y{-0.5} 
\edef\r{1.25}

@@

\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{SurvPlotNData.pdf}
\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells. The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{survprobwapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations. For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-3},\mu_w=1,\mu_a=1,\mu_m=1.$}
\label{SurvPlotNData}
\end{figure}

@@

\bibliographystyle{unsrt}
\bibliography{green2021}
\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\draw [postaction={decorate}, thick](-0.25,-4) -- (-0.25+5,0) node [midway, above] {$u$};
%\draw [postaction={decorate}, thick](-0.25,-2) -- (-0.25+5,-4) node [midway, above] {$2p$};
\draw [postaction={decorate}, thick](-0.25+5,-0) -- (-0.25+10,-4) node [midway, above] {$v$};
%\draw [postaction={decorate}, thick](-0.25+5+\r,-4) -- (-0.25+10-\r,-4) node [midway, above] {$u$};
%\draw [postaction={decorate}, thick](-0.25+5,-0) -- (-0.25+10,-4) node [midway, above] {$p$};
\draw [postaction={decorate}, thick]  (-0.25+10,-4) -- (-0.25+5,-8) node [midway, above] {$u$};
\draw [postaction={decorate}, thick](-0.25,-4)  -- (-0.25+5,-8)  node [midway, above] {$v$};

\filldraw[color=red!60, fill=red!5, very thick](-0.25,-4) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,0) circle (\r);
%\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,-4) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,-4) circle (\r);
%\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,-4) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,-8) circle (\r);

\draw [thick](0.05,-1.25+2-4) -- (0.05,-2.75+2-4);
\draw [thick](-0.55,-1.25+2-4) -- (-0.55,-2.75+2-4);

\draw [thick](0.15+5,-1.25+2) -- (0.15+5,-2.75+2);
\draw [thick](-0.65+5,-1.25+2) -- (-0.65+5,-2.75+2);
\draw [thick](-0.25+5,-1.25+2) -- (-0.25+5,-2.75+2);

\draw [thick](0.15+10,-1.25+2-4) -- (0.15+10,-2.75+2-4);
\draw [thick](-0.25+10,-1.25+2-4) -- (-0.25+10,-2.75+2-4);
\draw [,color=blue,thick](-0.7+10,-1.25+2-4) -- (-0.7+10,-2.75+2-4);

\draw [thick](0.05+5,-1.25+2-8) -- (0.05+5,-2.75+2-8);
\draw [,color=blue,thick](-0.55+5,-1.25+2-8) -- (-0.55+5,-2.75+2-8);
%\draw [thick](-0.55+5,-1.25-2) -- (-0.55+5,-2.75-2);

%\draw [thick](-0.55+10,-1.25-2) -- (-0.55+10,-2.75-2);
\end{tikzpicture}
\caption{
}
\label{figuretrisomy}
\end{figure}

\section{Model}
We model evolutionary rescue of cancer cells through aneuploidy with the help of multitype branching processes: we have three populations of cancer cells, wildtype euploid $w_t$, wildtype aneuploid $a_t$ and mutant euploid cells $m_t$ which proliferate and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively. Wildtype cells can acquire aneuploidy at rate $u$ while aneuploidy and wildtype cells can mutate with rate $v$ (se Figure \ref{figureAneuploidy}). Changes in the number of wildtype euploid, wildtype aneuploid and mutant euploid cells occur at rates:
\begin{subequations}
\begin{flalign}
w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
m_t&\rightarrow m_t-1:\quad \mu_am_t.
\end{flalign}
\end{subequations}
Given that we have used branching processes, the lineages produced by cells from the initial population will be independent of each other. This is not a biologically realistic model as cells compete for resources in tumors. We will study a more biologically realistic model in Section.
The survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation:
\begin{align}\nonumber
1-p_w&=\frac{\mu_w}{\lambda_w+\mu_w+u+v}+\frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right)+\frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2\\ \label{quadraticeqev1}
&+\frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right),
\end{align}
where the $p_a$ and $p_w$ are the survival probabilities of a population consisting initially of one aneuploidy and wildtype cell, respectively.

 The smallest solution of the quadratic equation \eqref{quadraticeqev1} gives the survival probability:
\begin{align}\label{survprobw}
p_w=\frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w}.
\end{align}
If the population originally consists of a single aneuploidy cell, then the probability that the population will survive is given by the following quadratic equation:
\begin{align}
1-p_a=\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2,
\end{align}
whose smallest solution is the survival probability:
\begin{align}\label{survproba}
p_a=\frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}.
\end{align}
The probability that a single mutant cell will survive is given by (see Appendix \ref{AppendixSurvLin}):
\begin{equation}
p_m=\left\{
  \begin{array}{@{}ll@{}}
  \frac{\lambda_m-\mu_m}{\lambda_m},\quad\text{if }\lambda_m>\mu_m \\
   0,\quad\text{else}.
  \end{array}\right.
  \end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\begin{tikzpicture}[very thick,decoration={
    markings,
    mark=at position 0.5 with {\arrow{>}}}, middlearrow/.style 2 args={
        decoration={             
            markings, 
            mark=at position 0.5 with {\arrow[xshift=3.333pt]{triangle 45}, \node[#1] {#2};}
        },
        postaction={decorate}
    }]
\edef\x{6.5} 
\edef\y{-0.5} 
\edef\r{1.25}

@@

\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{SurvPlotNData.pdf}
\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells. The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{survprobwapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations. For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
\label{SurvPlotNData}
\end{figure}

@@

\bibliographystyle{unsrt}
\bibliography{green2021}

\appendix

\section{Gillespie algorithm}
\begin{subequations}
\begin{flalign}
(+1,0,0)&:\quad \lambda_ww_t,\\
(-1,0,0)&:\quad \mu_ww_t,\\
(-1,+1,0)&:\quad uw_t,\\
(-1,0,+1)&:\quad vw_t,\\
(0,+1,0)&:\quad \lambda_aa_t,\\
(0,-1,0)&:\quad \mu_aa_t,\\
(0,-1,+1)&:\quad va_t,\\
(0,0,+1)&:\quad \lambda_am_t,\\
(0,0,-1)&:\quad \mu_am_t.
\end{flalign}
\end{subequations}

\section{Survival probability of a mutant lineage}\label{AppendixSurvLin}
\begin{align*}
p_{i+j,i}\left(\Delta t\right)=\left\{
  \begin{array}{@{}ll@{}}
  \mu i\Delta t+o\left(\Delta t\right),\quad j=-1 \\
  \lambda i\Delta t+o\left(\Delta t\right),\quad j=1 \\
    1-\left(\lambda+\mu\right) i\Delta t+o\left(\Delta t\right),\quad j=0 \\
   o\left(\Delta t\right),\quad j\neq-1,0,1.
  \end{array}\right.
\end{align*}

\begin{align*}
\frac{dp_i\left(t\right)}{dt}&=\lambda\left(i-1\right)p_{i-1}\left(t\right)+\mu\left(i+1\right)p_{i+1}\left(t\right)-\left(\lambda+\mu\right)ip_i\left(t\right)\\
\frac{dp_0\left(t\right)}{dt}&=\mu p_1\left(t\right)
\end{align*}

\begin{align*}
\mathcal{P}\left(z,t\right)=\left\{
  \begin{array}{@{}ll@{}}
  \left(\frac{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\mu\left(z-1\right)}{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\lambda\left(z-1\right)}\right)^N,\quad \text{if }\lambda\neq\mu \\
  \left(\frac{1-\left(\lambda t-1\right)\left(z-1\right)}{1-\lambda t\left(z-1\right)}\right)^N,\quad \text{if }\lambda=\mu
  \end{array}\right.
\end{align*}

\begin{equation*}
\mathcal{P}\left(z,t\right)=\sum_{i=0}^\infty p_i\left(t\right)z^i \quad\text{and}\quad p_i\left(t\right)=\left.\frac{1}{i!}\frac{\partial^i\mathcal{P}}{\partial z^i}\right|_{z=0}
\end{equation*}

\begin{align*}
p_0\left(t\right)=\left\{
  \begin{array}{@{}ll@{}}
  \left(\frac{\mu-\mu\e^{\left(\mu-\lambda\right)t}}{\lambda-\mu\e^{\left(\mu-\lambda\right)t}}\right)^N,\quad \text{if }\lambda\neq\mu \\
  \left(\frac{\lambda t}{1+\lambda t}\right)^N,\quad \text{if }\lambda=\mu
  \end{array}\right.
\end{align*}

\begin{align*}
p_0\left(\infty\right)=\lim_{t\rightarrow\infty}p_0\left(t\right)=\left\{
  \begin{array}{@{}ll@{}}
  1,\quad \text{if }\lambda\leq\mu \\
  \left(\frac{\mu}{\lambda}\right)^N,\quad \text{if }\lambda>\mu
  \end{array}\right.
\end{align*}
\end{document}
